\section{Open Problems}%
\label{sec:open_problems}
\todo[inline,color=red!80!green!25]{Citations}
There are many interesting directions to pursue in this area.
Below, we provide a few examples of random objects that may admit local access implementations.

\subparagraph*{Small Description Size}
\label{par:small_description_size}
\begin{itemize}
    \item Provide a local access implementation of degree queries for undirected random graphs, even for $G(n,p)$.
    How about $i^{th}$ neighbor queries?
    \item For simple models such as $G(n,p)$, provide a local access implementation of a \func{Random-Triangle}$(v)$ query,
    that returns a uniformly random triangle containing vertex $v$.
    \item Provide a memory-less local access implementation of basic queries for undirected random graphs.
    \item Given an ordered graph such as a lattice, provide an implementation to locally access a random perfect matching.
    Interesting special cases of this problem include random domino and lozenge tilings.
\end{itemize}

\subparagraph*{Huge Description Size}
\label{par:huge_description_size}
\begin{itemize}
    \item Provide a faster local access implementation for sampling the color of a specified vertex
    in a random $q$-coloring of a bounded degree graph $G$.
    \item Improve the local access implementation for sampling the color of a specified vertex in a random $q$-coloring of $G$,
    by supporting smaller values of $q$ (smaller than $9\Delta$).
    We remark that this problem in particular should be feasible, by simulating a faster mixing Markov chain.
    The important question is whether you can you get down to $q = 2\Delta$?
    \item Given query access to an input graph $G$ and starting vertex $v$,
    provide a local access implementation for sampling the location of a random walk starting at $v$ after $t$ steps.
    This may be feasible in certain restricted classes of graphs.
    \item Given query access to an input DNF formula, provide an implementation to access the truth value of a single variable in a uniformly random satisfying assignment.
\end{itemize}
