\section{Introduction}

The problem of computing local information of huge random objects
was pioneered in \cite{huge_old,huge}. 
Further work of \cite{sparse} considers the generation of sparse random $G(n,p)$ graphs
from the Erd\"{o}s-R\'{e}nyi model \cite{er}, with $p = O(\poly(\log n)/n)$,
which answers $\poly(\log n)$ \func{all-neighbors} queries,
listing the neighbors of queried vertices.
While these generators use polylogarithmic resources over their entire execution, 
they generate graphs that are  only guaranteed to {\em appear random} to algorithms
that inspect a {\em limited portion} of the generated graph.

In \cite{reut}, the authors construct an oracle for the generation of recursive trees,
and BA preferential attachment graphs.
Unlike \cite{sparse}, their implementation allows for an arbitrary number of queries.
This result is particularly interesting --  
although the graphs in this model are generated via a sequential process,
the oracle is able to locally generate arbitrary portions of it
and answer queries in polylogarithmic time.
Though preferential attachment graphs are sparse,
they contain vertices of high degree,
thus \cite{reut} provides access to 
the adjacency list through \func{next-neighbor} queries.
%On a high level, this is due to the fact that the BA model is directed, and all edges are independent.

In this work, we begin by \emph{formalizing} a model of local-access generators
implicitly used in \cite{reut}.
We next construct oracles that allow queries to both the adjacency matrix
and adjacency list representation of a basic class of random
graph families, without generating the entire graph at the onset:
Our oracles
provide \func{vertex-pair}, \func{next-neighbor}, and \func{random-neighbor} queries
for undirected graphs with {\em independent edge probabilities}, that is,
when each edge is chosen as an independent Bernoulli random variable. {\color{red} should really define what vertex-pair does explicitly -- readers might not be too familiar with it; the rest are intuitive enough; consider this in abstract as well.}
Using this framework, we construct the first \emph{efficient} local-access generators
for the Erd\"{o}s-R\'{e}nyi $G(n,p)$ model for all values of $p$, and the Stochastic Block model.
As in \cite{reut} (and unlike the generators in \cite{huge_old,huge,sparse}), 
our techniques allow unlimited queries.%(i.e. the entire graph can be generated).

The queries that our generator supports includes the queries supported by the generators
in all the previous works of \cite{reut, huge_old, huge, sparse},
   namely \func{vertex-pair}, \func{next-neighbor}, \func{all-neighbor}
(the latter only when the degree is small).
In addition, we provide the first implementation (to the best of our knowledge)
of \func{random-neighbor} queries, which do not follow trivially from the
\func{all-neighbor} queries in \emph{non-sparse graphs}.  For example,
such queries are useful in algorithms that employ random walks.
\func{random-neighbor} queries
present particularly interesting challenges,  since as we note in 
Section~\ref{par:random_neighbor_queries},
(1) \func{random-neighbor} queries affect the conditional probabilities
of the remaining neighbors in a non-trivial manner, and
(2) our implementation does not need to sample the degree of any vertex.
Our generator can support all {\color{red}four query types -- but for \func{all-neighbor} needs to assume poly-log degree;}
{\color{red} for undirected graphs with independent edge probabilities -- focus for this paragraph so might still worth clarifying somewhere?} %.
using $\mathcal{O}(\poly(\log n))$ time, space, and random bits
per query under assumptions on the ability to compute certain values
pertaining to consecutive edge probabilities. % (Section~\ref{sec:assumptions})
{\color{red} reminder: should also works for hyperbolic graphs, after proving that buckets are mostly non-empty; \func{random-neighbor} makes more sense than \func{next-neighbor} actually}

We then consider local-access generators for graphs in Kleinberg's Small World model.
In this case, the probabilities are based on distances in a 2-dimensional grid.
Using  a modified version of our previous sampling procedure,
we present such a generator supporting \func{all-neighbor} queries in 
$\mathcal{O}(\poly(\log n))$ time, space and random bits per query (since such graphs
are sparse, the other queries follow directly).
%by calculating CDFs efficiently. \textcolor{blue}{??? what do we mean by the last line?}

For additional related work, see Section~\ref{sec:related_work}. {\color{red} might still be good to leave in the appendix actually.}

%\input{introduction/results}
\input{introduction/new_results}
\input{introduction/related}
